---
title: "Stimuli selection from SPP"
author: '[Guillermo Montero-Melis](http://www.biling.su.se/montero_melis_guillermo)'
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
output:
  html_document:
    depth: 2
    number_sections: yes
    theme: default
    toc: yes
---


```{r setup, include=FALSE}
library("knitr")
library("ggplot2")
library("magrittr")
library("dplyr")
library("tidyr")
library("purrr")
# library(lme4)
# library(GGally)
knitr::opts_chunk$set(echo = TRUE)
```


Introduction
===========


Selection of critical arm/leg stimuli: English
====================================

We will do the whole process of stimuli selection very explicitly for English
and then we will just repeat the process (more concisely) for Swedish.


Critical verbs
-------

The English arm/leg verbs were normed by Swedish native speakers of L2 English.
Based on norming, we selected 80 verbs (40 arm-, 40 leg-related). Here are the
first ones:

```{r}
# Read data from disk
arm_leg <- read.csv("development/english-targets_with-info.csv", 
                    fileEncoding = "UTF-8", stringsAsFactors = FALSE) %>%
  rename(v_type = verb_category, score = Score, cogn_bin = cognate_bin)
# Add column that shows absolute bias
arm_leg$bias_abs <- abs(arm_leg$bias)
arm_leg %>% tail() %>% kable(digits = 2)
```

Columns: 

- `cogn_bin` indicates whether the English verb is a cognate of a Swedish verb,
and `cognate` shows that cognate if that's the case.
- `score` shows the proportion of correct translations of the English verb to
Swedish (by Swedish native speakers).
- `bias` shows the strength of the verb's bias towards arm- vs leg-relatedness.
Maximal arm bias is 6 and maximal leg bias is -6. A value of 0 would indicate
no bias.
- `bias_abs` shows the absolute value of `bias` (so we can easily compare 
strength of bias for arm and leg verbs).


Take a look at the bias per verb type:

```{r}
ggplot(arm_leg, aes(x = v_type, y = bias_abs)) + geom_boxplot()
```


What we need
------------

For the semantic priming task, we need:

1) 12 arm verbs that will serve as targets in the LDT
2) 12 arm verbs that will serve as primes, either related (if they prime arm
targets) or unrelated (if they prime leg targets), counterbalanced across two
lists
3) 12 arm verbs that serve as primes to non-words
4) We also need 1--3 for leg verbs.

We will order the verbs with regard to their bias, and take the 24 verbs with
the strongest bias of each type as items for 1) and 2) above. The 12 verbs per
category for 3) will be taken from the remaining verbs.

Note that we need to pair 12 arm/leg verbs with nonwords (3 above). Otherwise
participants could rapidly pick up that target primes (all of which they will 
already have encountered in the memory task) will always be followed by actual
verbs.



```{r}
# Lists of arm and leg words arranged according to bias and translation score
arm_leg_ord <- arm_leg %>% 
  split(.$v_type) %>% 
  map(~ arrange(.x, desc(.$bias_abs), desc(.$score))) %>%
  map("verb")
# Show first items in each list
arm_leg_ord %>% map(head)
```


Selection and cast into appropriate format
--------------------------------------

Create a function that takes the ordered vector of target items (arm or leg
verbs) and creates a list that contains:

- A dataframe of related pairs (e.g., arm-arm pairs) 
- A dataframe of critical primes paired with nonwords (e.g., arm-nonword pairs)

```{r}
mypairs <- function (vbs, verb_type) {
  # Select and shuffle 24 verbs with strongest bias
  strong_bias <- vbs[1 : 24] %>% base::sample()
  rel_df <- dplyr::tibble(
    condition = "related",
    item_type = paste(verb_type, verb_type, sep = "-"),
    prime  = strong_bias[1  : 12],
    target = strong_bias[13 : 24]
  )
  # Nonword pairs with placeholders for nonwords
  nonw_df <- dplyr::tibble(
    condition = "nonword",
    item_type = paste(verb_type, "NW", sep = "-"),
    prime  = base::sample(vbs[25 : length(vbs)], 12),
    target = "NW"
  )
  dplyr::bind_rows(rel_df, nonw_df)
}
```


```{r}
# Create dataframe with related pairs and nonword pairs (where the prime is
# an arm/leg word)
# Make use of imap to access the name of each list using .y shortcut
set.seed(555)
stim_relpairs_df <- imap(arm_leg_ord, ~ mypairs(.x, verb_type = .y)) %>%
  map_df(bind_rows)
stim_relpairs_df
```

Now we need to add an *unrelated* condition by pairing arm primes with leg
targets and vice versa:

```{r}
# Function to create UNRELATED items and combine with related and nonwords
# Note we want to implement counterbalancing (see McNamara, 2005, ch.8)
create_unrel <- function (df) {  # df is going to be stim_relpairs_df
  # the same nonword items appear in lists 1 and 2, so duplicate them...
  nonw <- df %>% dplyr::filter(condition == "nonword") %>% dplyr::bind_rows(., .)
  nonw$list <- rep(1:2, each = nrow(nonw) / 2)  # ...and assign them to each list
  # related items from which to create the unrelated ones:
  rel <- df %>% dplyr::filter(condition == "related")
  # Assign item pairs to 2 lists so half of each item_type is in each list:
  rel$list <- rep(1:2, each = nrow(rel) / 4)
  # Now split df into 4 nested lists (2 > 2) according to List and item_type
  nested <- rel %>% split(.$list) %>% map(~ split(x = ., f = .$item_type))
  
  # function to pair arm targets with leg primes and viceversa
  pair_unrel <- function (l_df) {  # l_df is each list of 2 dataframes in "nested"
    unrel <- tibble(
      condition = "unrelated",
      # The following assumes same number of arm-arm and leg-leg items:
      item_type = c(rep(c("arm-leg", "leg-arm"), each = nrow(l_df[["arm-arm"]]))),
      prime     = c(l_df[["arm-arm"]][["prime"]], l_df[["leg-leg"]][["prime"]]),
      target    = c(l_df[["leg-leg"]][["target"]], l_df[["arm-arm"]][["target"]])
    )
  }
  # unrelated pairs
  unrel <- nested %>% map(pair_unrel) %>% bind_rows()
  # Unrelated pairs go in opposite lists of the related pairs they were created from
  unrel$list <- rep(2:1, each = nrow(rel) / 2)
  # out
  bind_rows(nonw, rel, unrel) %>% dplyr::arrange(condition, item_type, list)
}
set.seed(7777)
stim_df <- create_unrel(stim_relpairs_df)
# Check out the result (and save to disk)
stim_df %T>%
  write.csv("development/check-out_stimuli.csv", fileEncoding = "UTF-8",
            row.names = FALSE) %>%
  kable()
```

Note that the `List` column has value NA for the nonword condition -- this 
actually means that all speakers see the same nonword items.



A few sanity checks
-------------------

The cross-tables below carry out a number of sanity checks for the *word*
item pairs (related or unrelated), and then for the *nonword* item pairs.

For targets and primes, they allow us to check that:

- In both lists, there are 12 related and 12 unrelated items
- Each target/prime is encountered once and once only per list
- Each target/prime appears once in the related and once in the unrelated 
condition

**Targets**

```{r}
with(stim_df %>% filter(condition != "nonword"),
     addmargins(table(condition, target, list)))
```

**Primes**

```{r}
with(stim_df %>% filter(condition != "nonword"),
     addmargins(table(condition, prime, list)))
```


**Nonword items**

```{r}
with(stim_df %>% filter(condition == "nonword"), addmargins(table(prime)))
```


**All conditions together**

```{r}
with(stim_df, addmargins(table(condition, prime, list)))
```






Add non-critical primes from SPP
==================================

We want to add the same amount of non-critical items as there are critical 
items to the semantic priming task: 24 target words that appear in the related
and unrelated conditions (counterbalancd across lists). We choose those items
from the [Semantic Priming Project](http://spp.montana.edu/), selecting items
that elicited a large priming effect at 200 ms SOA (see Hutchison et al.,
2013 in *Behav Res Meth* for details).


SPP database
------------

```{r}
# Load data I've previously downloaded from SPP website
spp_full <- read.csv("development/spp_assoc-related_181205.csv",
                     stringsAsFactors = FALSE) 
# Correct some typos in the data coding
# Relation 1
# levels(spp_full$Relation1)  # see typos
# table(spp_full$Relation1)   # see typos
rel1 <- as.character(spp_full$Relation1)
rel1 <- sub("antonymn", "antonym", rel1)
rel1 <- sub("unclassifed", "unclassified", rel1)
rel1 <- sub("Instrument", "instrument", rel1)
spp_full$Relation1 <- factor(rel1)
# Simplify data frame
spp <- spp_full %>%
  select(Target = TargetWord, Prime, Relation = Relation1, RT = LDT.200ms.RT,
         Acc = LDT.200ms.Acc, Priming_eff = LDT.200ms.RT.Priming,
         Priming_eff_z = LDT.200ms.Z.Priming)
spp %>% head() %>% kable(digits = 3)
```

Note that, according to the authors, "the standardized item score [`RT_z` and
`Priming_eff_z`] is the most accurate measure, minimizing the influence of a 
subject’s processing speed and variability" (Hutchison et al. 2013, p.1108).
Therefore, we focus on this measure.

Check correlations between raw and standardized measures:

```{r}
pairs(spp[, c("RT", "Priming_eff", "Priming_eff_z")])
```


And we are especially interested in the priming effects:


```{r}
ggplot(spp, aes(x = Priming_eff, y = Priming_eff_z)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm") +
  annotate("text", x = -200, y = 1.5, size = 10, parse = TRUE,
           label = paste("italic(r) == ", 
                         cor(spp$Priming_eff, spp$Priming_eff_z) %>% round(2))
           )
```



Criteria for item selection
---------------------------

We want to select targets that fulfil certain criteria:

1. Targets elicit a strong priming effect
2. Targets and primes -- both the related and unrelated prime -- are frequent
enough to be known by L2 speakers (use a sensible threhshold)
3. Avoid certain types of semantic relation that could be of a more idiomatic
kind, like forward/backward phrasal associates. Best if this can be somewaht
homogeneous.
4. Avoid prime-target pairs that conceptually overlap with critical pairs (e.g.,
PUSH-shove)
5. Avoid idiosyncratic items (e.g., proper names: JESUS-christ, ENGLAND-britain)


Load data files with relevant info and join with `spp` data frame:

```{r}
spp %>% head %>% kable
# Targets
spp_targ <- read.csv("development/spp_targets_181213.csv", stringsAsFactors = FALSE) %>%
  select(Target = TargetWord, tg_SubFreq = SubFreq, tg_LogSubFreq = LogSubFreq,
         tg_POS = POS)
spp_targ %>% head %>% kable
```

```{r}
# Related primes
spp_rel <- read.csv("development/spp_assoc-related_181205.csv", stringsAsFactors = FALSE) %>%
  select(Target = TargetWord, Prime, rel_SubFreq = SubFreq,
         rel_LogSubFreq = LogSubFreq, rel_POS = POS, rel_ELP_LDT_Acc = ELP.LDT.Acc)
spp_rel %>% head %>% kable
```


```{r}
# Unrelated primes
spp_unrel <- read.csv("development/spp_assoc-unrelated_181213.csv", stringsAsFactors = FALSE) %>%
  select(Target = TargetWord, Unrel = Unrelated, unrel_Acc = LDT.200ms.Acc)
spp_unrel %>% head %>% kable
```

Join them all:

```{r}
spp %<>% left_join(spp_targ) %>% left_join(spp_rel) %>% left_join(spp_unrel) %>%
   # reorder columns:
  select(Target : Relation, Unrel, RT, Acc, Priming_eff : rel_POS)
# Save for inspection in other scripts
spp %>% write.csv("development/spp_targets-primes-unrel.csv", row.names = FALSE)
head(spp, 3)
```


Have a look at frequency of targets and primes:

```{r, warning=FALSE, message=FALSE}
for (myfreq in c("tg_SubFreq", "tg_LogSubFreq", "rel_SubFreq", "rel_LogSubFreq")) {
  print(ggplot(spp, aes_string(x = myfreq)) + geom_histogram() + ggtitle(myfreq))
}
```

The raw counts are pretty useless, e.g.:

```{r}
summary(spp$tg_SubFreq)
# Remove them to gain some space when printing the object
spp$tg_SubFreq <- NULL
spp$rel_SubFreq <- NULL
```


So let's instead rely on the logged frequency (`tg_LogSubFreq` and 
`rel_LogSubFreq`) and select items whose logFreq is at least the median:

```{r}
spp_sel <- spp %>% 
  filter(tg_LogSubFreq >= median(tg_LogSubFreq, na.rm = TRUE),
         rel_LogSubFreq >= median(rel_LogSubFreq, na.rm = TRUE))
```



Items with strongest priming effect
----------------------------------

Here are the 50 items that elicited the strongest priming effect:


```{r}
spp_sel %>% top_n(50, wt = Priming_eff) %>% arrange(desc(Priming_eff)) %>%
  kable(digits = 2)
```


Add stop lists to exclude certain items:

```{r}
# Based on category: remove backward/forward phrasal association and items
# with no category assigned
stop_list_categ <- c("bpa", "fpa", "")
# Based on Target:
stop_list_target <- c(
  "christ", "jane", "jesus", "push"
)
```

```{r}
spp_sel %<>% filter(! ( Relation %in% stop_list_categ | Target %in% stop_list_target ))
spp_sel %>% top_n(50, wt = Priming_eff) %>% arrange(desc(Priming_eff)) %>%
  kable(digits = 2)
```


